A family of accessibility measures derived from spatial interaction principles

### 2025-05-21 paper/manuscript.qmd
added email to abstract -- for preprint

---
title: "A family of accessibility measures derived from spatial interaction principles"
subtitle: "Submitted preprint for peer-review (May 2025)"
author:
  - name: Anastasia Soukhov 
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Rafael H. M. Pereira
    email: rafael.pereira@ipea.gov.br
    affiliations:
        - name: Institute for Applied Economic Research - Ipea
          city: Brasília, Brazil
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliations:
        - name: University of Toronto Scarborough
          department: Department of Human Geography
          city: Toronto, Canada
    attributes:
        corresponding: true
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format:
  pdf:
    keep-tex: true
    documentclass: article
    header-includes: |
      \usepackage{draftwatermark} 
      \SetWatermarkText{DRAFT}
      \SetWatermarkScale{1.5}
      \SetWatermarkColor[gray]{0.95}
      \usepackage{float}
      \floatplacement{table}{H}
      \floatplacement{figure}{H}
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
#includes:
#      in_header: 'gt_packages.sty'
      #\usepackage{float}

@@

@@

In this section, we introduce a family of accessibility measures inspired by Wilson's spatial interaction modelling framework. At the end of this section, Table 8 presents a summary, where each member of the family of accessibility measure is explained in plain language, alongside their balancing factor, proportional allocation factor and mathematical equation and value interpretations.

>>>>

---
title: "A family of accessibility measures derived from spatial interaction principles"
subtitle: "Submitted preprint for peer-review (May 2025)"
author:
  - name: Anastasia Soukhov*
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Rafael H. M. Pereira
    email: rafael.pereira@ipea.gov.br
    affiliations:
        - name: Institute for Applied Economic Research - Ipea
          city: Brasília, Brazil
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliations:
        - name: University of Toronto Scarborough
          department: Department of Human Geography
          city: Toronto, Canada
    attributes:
        corresponding: true
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format:
  pdf:
    keep-tex: true
    documentclass: article
    header-includes: |
      \usepackage{draftwatermark} 
      \SetWatermarkText{DRAFT}
      \SetWatermarkScale{1.5}
      \SetWatermarkColor[gray]{0.95}
      \usepackage{float}
      \floatplacement{table}{H}
      \floatplacement{figure}{H}
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
#includes:
#      in_header: 'gt_packages.sty'
      #\usepackage{float}

@@

  *Email of corresponding author for preprint: soukhoa@mcmaster.ca

@@

In this section, we introduce a family of accessibility measures inspired by Wilson's spatial interaction modelling framework. At the end of this section, Table 1 presents a summary, where each member of the family of accessibility measure is explained in plain language, alongside their balancing factor, proportional allocation factor and mathematical equation and value interpretations.

### 2025-05-21 paper/manuscript.tex
added email to abstract -- for preprint

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A family of accessibility measures derived from spatial interaction principles},
  pdfauthor={Anastasia Soukhov; Rafael H. M. Pereira; Christopher D. Higgins; Antonio Páez},
  pdfkeywords={accessibility; spatial interaction model; balancing
factors; competitive accessibility; constraints; singly constrained;
doubly constrained; total constrained},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A family of accessibility measures derived from spatial
interaction principles}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Submitted preprint for peer-review (May 2025)}
\author{Anastasia Soukhov \and Rafael H. M. Pereira \and Christopher D.
Higgins \and Antonio Páez}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Transportation planning has long prioritized the efficiency of movement,
or mobility. However, the concept of accessibility represents a more
comprehensive evolution, shifting focus from mere movement to the
potential to reach (i.e., spatially interact) with desired destinations.
Despite growing recognition of accessibility-based planning approaches,
the concept remains fragmented, with inconsistent definitions and
unclear interpretations. This work's aim is to clarify and unify the
concept of accessibility by connecting it into spatial interaction
modeling. We demonstrate that widely used mobility and accessibility
models, such as gravity-based accessibility and spatial interaction
models, share common theoretical roots. From this foundation, this paper
offers three contributions: (A) we introduce a family of accessibility
measures within the principles of spatial interaction, and (B) formally
define four members of the family, namely the `unconstrained' measure
(i.e., Hansen-type accessibility), the `total constrained' measure
(i.e., a constrained version of the Hansen-type accessibility), the
`singly constrained' measure (i.e., related to the popular two step
floating catchment approach - 2SFCA), and the `doubly constrained'
measure representing realized interactions or `access', effectively
equal to the doubly constrained spatial interaction model; and (C) we
demonstrate the interpretability advantages of the family, as these
constrained accessibility measures yield values in units of the number
of potential ``opportunities for spatial interaction'' or ``population
for spatial interaction'' for each zone and zonal flow. The family of
accessibility measures proposed here clarifies the concept of
`potential' in accessibility, demonstrates theoretical and formulaic
linkages across popular accessibility and spatial interaction models,
and reintroduces measurement units into accessibility measures. By doing
so, we believe this family of measures can unlock a clearer, more
interpretable, and cohesive foundation for accessibility analysis.
\end{abstract}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A family of accessibility measures derived from spatial interaction principles},
  pdfauthor={Anastasia Soukhov*; Rafael H. M. Pereira; Christopher D. Higgins; Antonio Páez},
  pdfkeywords={accessibility; spatial interaction model; balancing
factors; competitive accessibility; constraints; singly constrained;
doubly constrained; total constrained},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A family of accessibility measures derived from spatial
interaction principles}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Submitted preprint for peer-review (May 2025)}
\author{Anastasia Soukhov* \and Rafael H. M. Pereira \and Christopher D.
Higgins \and Antonio Páez}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Transportation planning has long prioritized the efficiency of movement,
or mobility. However, the concept of accessibility represents a more
comprehensive evolution, shifting focus from mere movement to the
potential to reach (i.e., spatially interact) with desired destinations.
Despite growing recognition of accessibility-based planning approaches,
the concept remains fragmented, with inconsistent definitions and
unclear interpretations. This work's aim is to clarify and unify the
concept of accessibility by connecting it into spatial interaction
modeling. We demonstrate that widely used mobility and accessibility
models, such as gravity-based accessibility and spatial interaction
models, share common theoretical roots. From this foundation, this paper
offers three contributions: (A) we introduce a family of accessibility
measures within the principles of spatial interaction, and (B) formally
define four members of the family, namely the `unconstrained' measure
(i.e., Hansen-type accessibility), the `total constrained' measure
(i.e., a constrained version of the Hansen-type accessibility), the
`singly constrained' measure (i.e., related to the popular two step
floating catchment approach - 2SFCA), and the `doubly constrained'
measure representing realized interactions or `access', effectively
equal to the doubly constrained spatial interaction model; and (C) we
demonstrate the interpretability advantages of the family, as these
constrained accessibility measures yield values in units of the number
of potential ``opportunities for spatial interaction'' or ``population
for spatial interaction'' for each zone and zonal flow. The family of
accessibility measures proposed here clarifies the concept of
`potential' in accessibility, demonstrates theoretical and formulaic
linkages across popular accessibility and spatial interaction models,
and reintroduces measurement units into accessibility measures. By doing
so, we believe this family of measures can unlock a clearer, more
interpretable, and cohesive foundation for accessibility analysis.

*Email of corresponding author for preprint: soukhoa@mcmaster.ca
\end{abstract}
